---
title: 'Inference for numerical data'
author: "Your name goes here"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data}
library(mosaic)
library(tidyverse)
library(skimr)

nc <- read_csv(here::here("data", "nc.csv"))
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` or, even better, `skimr::skim()` command:

```{r summary}
summary(nc)
skim(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

Alternatively, you can use mosaic's `favstats` to get summary statistics of weight by habit (smoker vs. non-smoker, using __favstats(weight~habit, data=nc)__ )


```{r by-means}

favstats(weight~habit, data=nc)
```

There is an observed difference of 0.31 (7.14-6.83), but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test, where we will see whether there is enough 'evidence' to reject the hypothesis that the mean difference s equal to zero.

## Inference

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length` or just looking at the favstats output and see what the value for __n__ is for each group.

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

Next, we employ mosaic's \t-test\ and \conf.int\ functions, to conduct a 
hypothesis test and construct confidence intervals. 

```{r t-test-inf-weight-habit-ht, eval=FALSE, tidy=FALSE}
t.test(weight~habit, data=nc)

```
Mosaic's \t-test\ run's a Welch two sample t-test, with the null hypothesis being that the means are equal, or that the true difference in means is equal to 0. 

Working from the end of the output upwards, we first see the sample mean estimates (7.14 and 6.83), and before that we see a 9% confidence interval for the likely mean difference in weight, being anywhere between 0.05 and 0.58. This interval does not contain zero, so we do seem to have some evidence that the difference in birth weight between mothers who smoke and those that don't is __statistically significant__. The reported p-value of 0.02 is also less than the cut-off of 0.05, so again we do have some evidence to start doubting our null hypothesis of zero difference.


5.  Construct and record a confidence interval for the difference between the weights of babies born to smoking and non-smoking mothers. The confidence interval was actually calculated in the t-test, but if you wanted just theinterval, you can use the `confint`

By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$), as non-smoker is the first level in in the factor variable \habit\.

```{r inf-weight-habit-ci, tidy=FALSE}
confint(t.test(weight~habit, data=nc))

 # t.test(weight~habit, data=nc) %>% confint() # or you can just pipe the t.test into confint()

```

* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) by habit and interpret it in context. Is the difference statistically singificant? Please explain.

-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the `t.test` function: `conf.level = 0.90`.

-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `t.test` 
function, report the statistical results, and also provide an explanation in 
plain language.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab 
written by the faculty and TAs of UCLA Statistics.
</div>